# Reproducible Research: Assignment 1


## Loading and preprocessing the data

```{r echo=TRUE}
originalData <- read.csv("activity.csv")
```

First 5 rows of original data:
```{r echo=FALSE}
print(originalData[1:5,])
```

## What is mean total number of steps taken per day?

Creating dataset containing the total number of steps taken each day

```{r echo=TRUE}
dailyStepSum <- aggregate(originalData$steps, list(originalData$date), sum)
```
First 5 rows of daily Step Sum data:
```{r echo=FALSE}
colnames(dailyStepSum) <- c("Date", "Steps")
print(dailyStepSum[1:5,])
```

2. Creating histogram of daily Step Data

```{r fig.width=15, echo=TRUE, tidy=FALSE}
with(dailyStepSum, {
        par(oma=c(2,0,0,0), mar=c(6.75,6.75,3,0), mgp=c(5.75,0.75,0), las=2)
        barplot(
                height=Steps,
                main="Total Steps taken per Day",
                xlab="Date",
                ylab="Step count per Day",
                names.arg=Date,
                space=c(0)
        )
})
```

3. Calculating the mean and median values (ignoring NA values) using the daily step count dataset.

1. Mean
```{r echo=TRUE}
dailyStepMean <- mean(dailyStepSum$Steps, na.rm=TRUE)
```
```{r echo=FALSE}
print(dailyStepMean)
```
2. Median
```{r echo=TRUE}
dailyStepMedian <- median(dailyStepSum$Steps, na.rm=TRUE)
```
```{r echo=FALSE}
print(dailyStepMedian)
```

## What is the average daily activity pattern?

Finding the mean (average) steps taken for eatch 5-minute time interval averaged over all the days in the data-

1. Generating the mean (average) number of steps taken (ignoring NA values) for each 5-minute interval, itself averaged across all days.

```{r echo=TRUE, tidy=FALSE}
intervalSteps <- aggregate(
        data=originalData,
        steps~interval,
        FUN=mean,
        na.action=na.omit
)
colnames(intervalSteps) <- c("Interval", "AvgStepsAvgAcrossDay")
```
First 5 rows in new dataset:
        ```{r echo=FALSE}
print(intervalSteps[1:5,])
```

2. A Time-Series plot is created from the above dataset

```{r fig.width=15, echo=TRUE, tidy=FALSE}
with(intervalSteps, {
        plot(
                x=Interval,
                y=AvgStepsAvgAcrossDay,
                type="l",
                main="Time-Series of Average StepCount against 5 min Interval",
                xlab="5-minute Interval",
                ylab="Average StepsCount (Averaged across all Days)"

        )
})
```

3. Finding the 5-minute interval with the maximum number of steps

```{r echo=TRUE, tidy=FALSE}
intervalMax <- intervalSteps[intervalSteps$AvgStepsAvgAcrossDay==max(intervalSteps$AvgStepsAvgAcrossDay),]
```
```{r echo=FALSE}
print(intervalMax)
```
Conclusion:

The interval between **`r as.character(intervalMax[1])`** and  **`r as.character(as.numeric(intervalMax[1])+5)`** minutes has the maximum number of steps.


## Imputing missing values

The goal of this section is to generate a new graph using the same data as from the first section but with its NA values replaced.

Let's use mean (average) 5-minunte interval values from the previous section to replace the NA values.

1. Total number of rows with NA values in original data.

```{r echo=TRUE}
countNA <- nrow(subset(originalData, is.na(originalData$steps)))
```
```{r echo=FALSE}
print(countNA)
```


Rounding decimal to a whole number.

```{r echo=TRUE, tidy=FALSE}
stepValues <- data.frame(originalData$steps)
stepValues[is.na(stepValues),] <- ceiling(tapply(X=originalData$steps,INDEX=originalData$interval,FUN=mean,na.rm=TRUE))
newData <- cbind(stepValues, originalData[,2:3])
colnames(newData) <- c("Steps", "Date", "Interval")
```

5 rows of new dataset:
```{r echo=FALSE}
print(newData[1:5,])
```

3. The total number of steps taken each day is generated using this new dataset.

```{r echo=TRUE}
newDailyStepSum <- aggregate(newData$Steps, list(newData$Date), sum)
```
5 rows of new dataset:
```{r echo=FALSE}
colnames(newDailyStepSum) <- c("Date", "Steps")
print(newDailyStepSum[1:5,])
```

4. A histogram of the above data is created as a form of visual representation.

```{r fig.width=15, echo=TRUE, tidy=FALSE}
with(newDailyStepSum, {
        par(oma=c(2,0,0,0), mar=c(6.75,6.75,3,0), mgp=c(5.75,0.75,0), las=2)
        barplot(
                height=Steps,
                main="Graph of StepsCount per Day",
                xlab="Dates",
                ylab="StepCount per Day",
                names.arg=Date,
                space=c(0)
        )
})
```

5. Calculate the mean and median values of this new dataset (NA values replaced with mean).

1. Mean
```{r echo=TRUE}
newDailyStepMean <- mean(newDailyStepSum$Steps)
```
```{r echo=FALSE}
print(newDailyStepMean)
```
2. Median
```{r echo=TRUE}
newDailyStepMedian <- median(newDailyStepSum$Steps)
```
```{r echo=FALSE}
print(newDailyStepMedian)
```

Observation :
Adding the missing values to the original data has caused both the mean and median values to increase.

1. Mean:
        `r as.character(floor(as.numeric(dailyStepMean)))` to `r as.character(floor(as.numeric(newDailyStepMean)))`

2. Median:
        `r as.character(floor(as.numeric(dailyStepMedian)))` to `r as.character(floor(as.numeric(newDailyStepMedian)))`


## Are there differences in activity patterns between weekdays and weekends?
>For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

        1.  A new column indicating whether the date is a weekday or a weekend is added to the new dataset created in the previous section.

```{r echo=TRUE}
dateDayType <- data.frame(sapply(X=newData$Date, FUN=function(day) {
        if (weekdays(as.Date(day)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {
                day <- "weekday"
        }
        else {
                day <- "weekend"
        }
}))

newDataWithDayType <- cbind(newData, dateDayType)

colnames(newDataWithDayType) <- c("Steps", "Date", "Interval", "DayType")
```

First 5 rows of new data:
        ```{r echo=FALSE}
print(newDataWithDayType[1:5,])
```
2. The data is then separated into weekday or weekend and the mean (average) number of steps taken for each 5-minute interval, itself averaged across all weekday days or weekend days is calculated.

```{r echo=TRUE, tidy=FALSE}
dayTypeIntervalSteps <- aggregate(
        data=newDataWithDayType,
        Steps ~ DayType + Interval,
        FUN=mean
)
```
First 5 rows of the dataset:
```{r echo=FALSE}
print(dayTypeIntervalSteps[1:5,])
```

3. Finally, a panel plot of both weekend and weekday graphs is generated.

```{r fig.width=15, echo=TRUE, tidy=FALSE}
library("lattice")

xyplot(
        type="l",
        data=dayTypeIntervalSteps,
        Steps ~ Interval | DayType,
        xlab="Interval",
        ylab="StepsCount",
        layout=c(1,2)
)
```
